Seminar: Programming models and Code Generation\\ Winter Term 2014/2015\\ {\bf Quasi deterministic parallel programming using LVars}

### 2015-01-03 draft.tex
new update

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{lstlisting}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
   _ = put lv 4  /write 4 in lv/
	  in v   /final value is v/
\end{lstlisting}
The \textit{let par} expression, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. \par 
\begin{lstlisting}
let par  _ = put l (4, )
         _ = put l (, 3) /error/
         in get l
\end{lstlisting}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

        \item \textit{Single-assignment parallelism}: IVars are an example which are memory locations that can be either full or empty and can only be assigned once. Reads are blocking for IVars. They have been used in the \textit{monad-par} Haskell library, \textit{Intel CnC(Concurrent collections, Concurrent ML(as SnycVars)}.,

@@

In a directed graph where v represents a particular node, we wish to find all (or the first k degrees of) node connected to v, then analyze each node in that set. Existing parallel solutions\cite{kuper} often use a non-deterministic traversal of the connected component even though the final connected component is deterministic. For example, IVars cannot accumulate sets of visited nodes, nor can they be used as “mark bits” on visited nodes, since they can only be written once and not tested for emptiness. Streams, on the other hand, impose an excessively strict ordering for computing the unordered set of vertex labels in a connected component. A purely functional attempt is given in the original paper\cite{lkuper} where connected-component discovery is parallel, but members of the output set do not become available to other computations until component discovery is finished, limiting parallelism. 

@@

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that a write to a location is the least upper bound of the new value and the current value. 
\begin{lstlisting}
let par _ = put num 3
        _ = put num 2
  in get num /always 3/
\end{lstlisting}
Monotonic writes alone aren't enough to ensure deterministic programs. The above program is non-deterministic where num's notion of “bigger” is “>=”. Despite writes always being monotonically increasing. Get might read either 2 or 3, depending on the order of $put or get$ operations determined by the scheduler. \par
\begin{itemize}
\item To maintain determinism, $get$ operation is restricted: instead of reading the LVar’s exact value, we read one of a set of lower bounds on its value that are given as an extra argument to $get$. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $Get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num's value it is at least 3, it will remain at or above 3 forever. As long as we only access them through the $put or get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

To describe Lvar formally, a parallel call-by-value $\lambda$-calculus named $\lambdaLVar$ is introduced in the paper\cite{lkuper} which is extended with a $store$ and communication primitives $put$ and $get$ that operate on data in the $store$. The definition of $\lambdaLVar$ is parameterized by the choice of lattice and therefore $\lambdaLVar$ is actually a family of languages, rather than a single language. An LVar may have an arbitrary number of states forming a set $D$, which is partially ordered by a relation $\userleq$. Formally, $D$ is a \emph{bounded-join-semilattice} augmented with a greatest element $\top$.

@@

During the evaluation of a $\lambdaLVar$ program, a \emph{store} $S$ keeps track of the states of LVars. Each LVar is represented by a \emph{binding} from a location $l$, drawn from a countable set $\Loc$, to its \emph{state}, which is some element $d \in D$. Although each LVar in a program has its own state, the states of all the LVars are drawn from the same lattice $D$. We can do this with no loss of generality because lattices corresponding to different types of LVars could always be union ed into a single lattice (with shared $\bot$ and $\top$ Elements). \\ \\

@@

The new, put, and get operations create, write to, and read from LVars, respectively.

@@

\item $new$ extends the store with a binding for a new LVar whose initial state is $\bot$, and returns the location $l$ of that LVar (i.e. a pointer to the LVar).

@@

\label{f:lvars-frame-rule}

@@

For Lambda Lvar syntax and semantics, remaining definitions and its details please refer\cite{lkuper}.\\ \\ 
\textbf{\textit{Proof of determinism:}} The key to the determinism is the frame rule given in \cite{Figure1}. A frame property captures the idea of local reasoning about programs that alter state\cite{lindsey}. Here, C is a program, and $\{p\} C \{q\}$ says that if the assertion p  is true before C  runs, then the assertion q  will be true afterwards. For example, $p$ and $q$  might describe the state of the heap before and after it is updated by $C$. It tells that running $C$ starting from a state satisfying the assertion $p$ $∗$ $r$ will result in a state satisfying the assertion $q$ $∗$ $r$. The assertion $p$ $*$ $r$ is satisfied by a heap if the heap can be split into two non-overlapping parts satisfying $p$ and $r$, respectively.

For LVars, the frame property says that independent effects commute with each other. Consider an expression $e$ that runs starting in \emph{store S} and steps to the expression $e′$, updating the \emph{store} to $S′$. Frame property guarantees that if $e$ starts from a larger $store S$ $\sqcup$ $R$, where $R$ is some other store \emph{“framing on”} to $S$, then $e$  will update the $store$ to $S′$ $\sqcup$ $R$, and $e$ will step to $e′$. The \emph{“frame”} \emph{store} $R$  is a \emph{store} resulting from some other independently-running computation. $S$ and $R$ (or $S′$ and $R$) need not be disjoint — we are combining them with $\sqcup$, not with $∗$. If they did have to be disjoint, it would mean that concurrent operations cannot touch the same parts of the store. For LVars, this kind of total disjointness is unnecessary, since updates commute! Instead, we can behave as though $S$ and $R$ are disjoint, even though they might not be. \\ \\
\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to\cite{lkuper}. Lemmas there also show how cases of deadlocks and livelocks are also handled.

@@

\begin{lstlisting}

@@

\end{lstlisting}

@@

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the runParThenFreeze function.

\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the original LVar formalism to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern—which is a simpler primitive combining all others.

>>>>

The purpose of a deterministic-by-construction programming model is to rule out programs like the following from being written, or force them to behave deterministically. \cite{lindsey}Written in a hypothetical parallel language. 
\begin{verbatim}
let _ = put lv 3 in  /write 3 in lv/
let par v = get lv  /read from lv/
        _ = put lv 4  /write 4 in lv/
	      in v   /final value is v/
\end{verbatim} 
The \textit{let par} expression which has \textit{fork-join semantics}, evaluates the two sub-expressions in parallel. The value of $v$ is the value of the entire program. Depending on the scheduler \textit{get} $lv$ or \textit{put} $lv\ 4$ can run first, $v$ might be either $3$ or $4$ during multiple runs giving rise to non determinism.\par
A popular approach for enforcing determinism is to require that variables can only be assigned to once, giving us a single-assignment language. Our program would be invalid because it tries to write to $lv$ twice. Single-assignment variables with blocking read semantics, often known as IVars, turn up in all kinds of deterministic parallel systems. But the single-assignment rule disallows a number of programs that we would like to permit such as writing the same value to an IVar twice. As another example, consider the problem of filling in various parts of a data structure in parallel. \cite{lindsey}Say that $lv$ contains a pair, and we are writing to $lv$ twice, with each write filling in one component of the pair. Please note that $x$ is equal to $\bot$ below.\par 
\begin{verbatim}
let par  _ = put lv (4,x)
         _ = put lv (x,3) /error/
         in get lv
\end{verbatim}  
Read from $lv$ is possible once both writes have completed, so it will always be (4, 3). But single-assignment rules out this program(although we could fake it by using a distinct IVar for each component of the pair, but with more sophisticated data structures it becomes difficult to fake a solution with IVars.)\par

@@

        \item \textit{Single-assignment parallelism}: IVars are an example which are memory locations that can be either full or empty and can only be assigned once. Reads are blocking for IVars. They have been used in the \textit{monad-par} Haskell library, \textit{Intel CnC(Concurrent collections), Concurrent ML(as SnycVars)}.,

@@

In a directed graph where v represents a particular node, we wish to find all (or the first k degrees of) node connected to v, then analyze each node in that set. Existing parallel solutions\cite{kuper} often use a non-deterministic traversal of the connected component even though the final connected component is deterministic and IVars and Streams are not of much help. For example, IVars cannot accumulate sets of visited nodes, nor can they be used as “mark bits” on visited nodes, since they can only be written once and not tested for emptiness. Streams, on the other hand, impose an excessively strict ordering for computing the unordered set of vertex labels in a connected component. A purely functional attempt is given in the original paper\cite{lkuper} where connected-component discovery is parallel, but members of the output set do not become available to other computations until component discovery is finished, limiting parallelism. 

@@

\section{LVars and LVish}
\subsection{LVars}
An LVar is a memory location that can be shared between multiple threads whose contents can only grow bigger over time, for some definition of “bigger” that the user of the LVar gets to specify. \par
LVars, short for lattice variables, are a generalization of IVars, which are variables that can only be written to once. (The “I” is for “immutable”.) LVars, on the other hand, allow multiple writes, as long as those writes are monotonically increasing with respect to a user-specified lattice. Meanwhile, LVar reads have to specify a threshold set of possible states. A read will block until one of the states in the threshold set has been reached, and then return that state. Information can never be removed and the order in which it is added is not observable. IVars turn out to be a special case of LVars. 
The property of monotonically increasing write operations is enforced by the language. The semantics of $put$ says that a write to a location is the least upper bound of the new value and the current value. 
\begin{verbatim}
let par _ = put num 3
        _ = put num 2
        x = get num
  in x  /*could be 2 or 3*/
\end{verbatim}
Monotonic writes alone aren't enough to ensure determinism. The above program is non-deterministic where num's notion of “bigger” is $\geq$. Despite writes always being monotonically increasing. $Get$ might read either $2$ or $3$, depending on the order of $put$ or $get$ operations determined by the scheduler. With an LVar model this behavior changes.\par
\begin{itemize}
\item To maintain determinism, the LVars model restricts the $get$ operation. Instead of reading the LVar’s exact value, it reads one of a set\{threshold set\} of lower bounds on its value that are given as an extra argument to $get$ not shown in the example. The $get$ operation will block until the LVar reaches a value that is at or above one of those lower bounds. For the above example, the threshold set will just be \{$3$\}. $Get$ operation is called a threshold read, because the value it returns is the threshold we have crossed.
\item Once num value is atleast $3$, it will remain at or above $3$ forever. As long as we only access them through the $put$ or $get$ interface, LVars are thread-safe: we can safely share them between threads without introducing unintentional non-determinism. That is, $puts$ and $gets$ can happen in any order, without changing the value that a program evaluates to.
\end{itemize}
%\begin{center}
%\textit{Monotonically increasing writes + threshold reads = deterministic parallelism}
%\end{center}
The resulting model is general enough to subsume the existing single-assignment model. The user-specified set just needs to be a \textit{join-semilattice}, which is a partially ordered set in which every two elements have a least upper bound.

To describe Lvar formally, a parallel call-by-value $\lambda$-calculus named $\lambdaLVar$ is introduced in \cite{lkuper} which is extended with a $store$ and communication primitives $put$ and $get$ that operate on data in the $store$. The definition of $\lambdaLVar$ is parameterized by the choice of lattice and therefore $\lambdaLVar$ is actually a family of languages, rather than a single language. An LVar may have an arbitrary number of states forming a set $D$, which is partially ordered by a relation $\userleq$. Formally, $D$ is a \emph{bounded-join-semilattice} augmented with a greatest element $\top$.

@@

During the evaluation of a $\lambdaLVar$ program, a \emph{store} $S$ keeps track of the states of LVars. Each LVar is represented by a \emph{binding} from a location \emph{$l$}, drawn from a countable set $\Loc$, to its \emph{state}, which is some element $d \in D$. Although each LVar in a program has its own state, the states of all the LVars are drawn from the same lattice $D$. We can do this with no loss of generality because lattices corresponding to different types of LVars could always be union ed into a single lattice (with shared $\bot$ and $\top$ Elements). \\ \\

@@

The $new$, $put$, and $get$ operations create, write to, and read from LVars, respectively.

@@

\item $new$ extends the $store$ with a binding for a new LVar whose initial state is $\bot$, and returns the location $l$ of that LVar (i.e. a pointer to the LVar).

@@

\label{f:frame-rule}

@@

For $\lambdaLVar$ syntax and semantics, remaining definitions and its details please refer\cite{lkuper}.\\ \\ 
\textbf{\textit{Proof of determinism:}} The key to the determinism is the frame rule given in \ref{f:frame-rule}. A frame property captures the idea of local reasoning about programs that alter state\cite{lindsey}. Here, $C$ is a program, and $\{p\} C \{q\}$ says that if the assertion $p$ is true before $C$ runs, then the assertion $q$ will be true afterwards. For example, $p$ and $q$  might describe the state of the heap before and after it is updated by $C$. It tells that running $C$ starting from a state satisfying the assertion $p$ $\ast$ $r$ will result in a state satisfying the assertion $q$ $\ast$ $r$. The assertion $p$ $\ast$ $r$ is satisfied by a heap if the heap can be split into two non-overlapping parts satisfying $p$ and $r$, respectively.

For LVars, the frame property says that independent effects commute with each other. Consider an expression $e$ that runs starting in \emph{store S} and steps to the expression $e{′}$, updating the \emph{store} to $S{′}$. Frame property guarantees that if $e$ starts from a larger $store S$ $\sqcup$ $R$, where $R$ is some other store \emph{“framing on”} to $S$, then $e$  will update the $store$ to $S′$ $\sqcup$ $R$, and $e$ will step to $e′$. The \emph{“frame”} \emph{store} $R$  is a \emph{store} resulting from some other independently-running computation. $S$ and $R$ (or $S′$ and $R$) need not be disjoint — we are combining them with $\sqcup$, not with $∗$. If they did have to be disjoint, it would mean that concurrent operations cannot touch the same parts of the store. For LVars, this kind of total disjointness is unnecessary, since updates commute! Instead, we can behave as though $S$ and $R$ are disjoint, even though they might not be. \\ \\
\textbf{\textit{Diamond Lemma:}} This does the heavy lifting of the determinism proof: it establishes the diamond property, which says that if a configuration steps to two different configurations, there exists a single third configuration to which those configurations both step. For supporting lemmas, its details and complete proofs please refer to\cite{lkuper}. Lemmas there also show how cases of deadlocks and livelocks are also handled.

@@

\begin{verbatim}

@@

\end{verbatim}

@@

Freezing introduces quasi-determinism to the programming model, because we might freeze before we have quiesced, or because our thread that's freezing and quiescing might be racing with some other thread to write into a shared LVar. But, if it is ensured that freezes only ever happen after all writes have completed, then the computations will be deterministic, because there wouldn't be any write-after-freeze races. This is enforced at the implementation level which takes care of quiescing and
freezing at the end of a computation. In the LVish library, this is provided by the runParThenFreeze function.\\
\cite{lkuper2}contains a quasi-deterministic, parallel, call-by-value $\lambdaLVish$-calculus extended with a store containing LVars. It extends the $\lambdaLVar$ calculus to support event handlers and freezing. Rather than modeling the full ensemble of event handlers, handler pools, quiescence, and freezing as separate primitives, it instead formalizes the \emph{“freeze-after”} pattern—which is a simpler primitive combining all others.
